---
title: "SIREN Cox models - R implementation"
author: 
- "Peter Kirwan, MRC Biostatistics Unit, University of Cambridge"
date:   "`r Sys.Date()`"
output: 
    html_document:
        code_folding: hide
        toc:          true
---

```{r setup, include=FALSE}

# libraries
library(tidyverse)
library(survival)
library(finalfit)
library(flextable)
library(patchwork)

# load data
load("~/coviddata/siren_post_processing.RData")

```

```{r labels}

# define labels
ve_label <- c(
    "Unvaccinated",
    "d1 0-20 PF",
    "d1 20+ PF",
    "d2 0-13 PF",
    "d2 14-73 PF",
    "d2 74-133 PF",
    "d2 134-193 PF",
    "d2 194-253 PF",
    "d2 254+ PF",
    "d3 0-6 3PF",
    "d3 7-66 3PF",
    "d3 67-126 3PF",
    "d3 127+ 3PF",
    "d2 0-13 AZ",
    "d2 14-73 AZ",
    "d2 74-133 AZ",
    "d2 134-193 AZ",
    "d2 194+ AZ",
    "d3 0-6 3AZ",
    "d3 7-66 3AZ",
    "d3 67+ 3AZ",
    "d3 0-6 2AZ/1PF",
    "d3 7-66 2AZ/1PF",
    "d3 67+ 2AZ/1PF"
)

pi_label <- c("Naive", "3-6 months","6-9 months","9-12 months","12-15 months","15+ months")

```

This document presents Cox proportional hazards estimates for the SIREN interim analysis dataset (SIREN_Interim3_20220228.dta). This analysis replicates a set of Stata analyses, thereby providing independent verification.

Survival analysis datasets with multiple intervals for each participant and corresponding covariate values were prepared using the `survival` package (_Therneau T_) and formatted using the `finalfit` package (_Harrison E, Drake T, and Ots R_).

Minor differences exist between the Stata and `survival` package implementations of the Cox Proportional Hazards model:

- The default method for handling tied event times in Stata is the Breslow approximation, whereas in the `survival` package the more accurate Efron approximation is the default. We decided to use the Efron approximation.

- Clustering on NHS trust is used for robust variance calculation. When using clustering, Stata makes an adjustment to the standard errors - multiplying by g/(g − 1), where g is the number of clusters - whereas the `survival` package does not make this adjustment. We decided to use the adjusted standard errors.

## 1. Alpha Delta period

Analysis for the pre-Omicron period (7th December 2020 - 30th November 2021). 

The vaccination and primary infection categories are shown below. Efficacy of prior infection with an alternate categorisation is also included. To assess protection following prior infection, individuals with a primary infection occurring prior to 1st September 2021 re-entered the dataset after 90 days.

### Data and proportional hazards test

```{r ve, fig.width=10, warning=FALSE, message=FALSE}

# define variables
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "vaccine_cat"
explanatory_multi = c("vaccine_cat", "primary_inf", "staff_type", "strata(agegr)", "strata(work_exposure_frequency)", "cluster(trust_code)")

# format data
model_data <- split_ve %>% 
    mutate(vaccine_cat=factor(vaccine_cat, labels = ve_label)) %>% 
    filter(eligible == 1,
           alpha_delta == 1,
           vaccine_cat %in% c("Unvaccinated",
                              "d2 0-13 PF",
                              "d2 14-73 PF",
                              "d2 74-133 PF",
                              "d2 134-193 PF",
                              "d2 194-253 PF",
                              "d2 254+ PF",
                              "d3 7-66 3PF",
                              "d2 14-73 AZ",
                              "d2 74-133 AZ",
                              "d2 134-193 AZ",
                              "d2 194+ AZ",
                              "d3 7-66 3AZ",
                              "d3 7-66 2AZ/1PF")) %>% 
    mutate(vaccine_cat = fct_collapse(vaccine_cat, "d2 194+ PF" = c("d2 194-253 PF", "d2 254+ PF")),
           vaccine_cat = fct_drop(vaccine_cat),
           primary_inf = factor(primary_inf, labels = pi_label)
           )

# check vaccine category groupings
model_data %>%
    group_by(vaccine_cat) %>%
    summarise(
        particiants = n_distinct(study_id),
        events = sum(event),
        exposure = sum(tstop - tstart)
    ) %>%
    flextable() %>% 
    autofit()

model_data %>%
    group_by(primary_inf) %>%
    summarise(
        particiants = n_distinct(study_id),
        events = sum(event),
        exposure = sum(tstop - tstart)
    ) %>%
    flextable() %>% 
    autofit()

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    labs(title = "Kaplan-Meier survival curve")+
    theme_minimal()

# test proportional hazards assumption
# not significant - indicates proportional hazards assumption holds
uni_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

uni_ph$table %>% knitr::kable(caption = "Proportional hazards for univariable model")

multi_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

multi_ph$table %>% knitr::kable(caption = "Proportional hazards for multivariable model")

# table, finalfit provides both unadjusted and adjusted hazards
# model_data %>% finalfit(dependent, explanatory_multi) %>% 
#     flextable() %>% 
#     autofit()

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```

### Hazard ratios from multivariable model

```{r ve-fig, fig.height=10, fig.width=15, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```

```{r efficacy}

cox_result <- model_data %>% coxphmulti(dependent, explanatory_multi)

# vaccination
# cox_result %>% 
#     broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
#     filter(grepl("vaccine", term)) %>% 
#     mutate(term = sub("vaccine_cat","",term),
#            term = factor(term, levels = c("Unvaccinated",
#                               "d2 0-13 PF",
#                               "d2 14-73 PF",
#                               "d2 74-133 PF",
#                               "d2 134-193 PF",
#                               "d2 194+ PF",
#                               "d3 7-66 3PF",
#                               "d2 14-73 AZ",
#                               "d2 74-133 AZ",
#                               "d2 134-193 AZ",
#                               "d2 194+ AZ",
#                               "d3 7-66 3AZ",
#                               "d3 7-66 2AZ/1PF"))) %>% 
#     ggplot() + 
#     aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high) + 
#     geom_point() +
#     geom_linerange() +
#     labs(x = "", y = "1 - HR", title = "Efficacy of vaccination against infection during Alpha/Delta period", subtitle = "Reference: Unvaccinated") +
#     coord_cartesian(ylim = c(0,1)) +
#     theme_minimal()+
#     theme(axis.text.x = element_text(angle = 45, hjust=1))

# prior infection
p2 <- cox_result %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
    filter(grepl("primary", term)) %>% 
    mutate(term = sub("primary_inf","",term),
           term = factor(term, levels = pi_label)) %>% 
    ggplot() + 
    aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high) + 
    geom_point() +
    geom_linerange() +
    labs(x = "", y = "1 - HR", subtitle = "A. Fine timescale") +
    coord_cartesian(ylim = c(0,1)) +
    theme_minimal()

model_data <- model_data %>% 
    mutate(primary_inf = fct_collapse(primary_inf, "<12 months" = c("3-6 months","6-9 months","9-12 months"),
                                                   "12+ months" = c("12-15 months","15+ months")))

# prior infection, 1 year cutoff
cox_result <- model_data %>% coxphmulti(dependent, explanatory_multi)

p3 <- cox_result %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
    filter(grepl("primary", term)) %>% 
    mutate(term = sub("primary_inf","",term)) %>% 
    ggplot() + 
    aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high) + 
    geom_point() +
    geom_linerange() +
    labs(x = "", y = "1 - HR", subtitle = "B. Course timescale") +
    coord_cartesian(ylim = c(0,1)) +
    theme_minimal()

p2 + p3 + plot_annotation(title = "Efficacy of prior infection against future infection during Alpha/Delta period", caption = "Reference group: Naive") + plot_layout(widths = c(2, 1))

```

## 2. Omicron period

Analysis for the Omicron period (10th December 2021 - 28th February 2022). 

The vaccination and primary infection categories are shown below. Efficacy of prior infection with an alternate categorisation is also included. To assess protection following prior infection, individuals with a primary infection occurring prior to 1st September 2021 re-entered the dataset after 90 days.

### Data and proportional hazards test

```{r ve-om, fig.width=10, warning=FALSE, message=FALSE}

# define variables
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "vaccine_cat"
explanatory_multi = c("vaccine_cat", "primary_inf", "staff_type", "gender", "ethngr", "strata(region)", "strata(agegr)", "strata(work_exposure_frequency)", "cluster(trust_code)")

# format data
model_data <- split_ve %>% 
    mutate(vaccine_cat=factor(vaccine_cat, labels = ve_label)
           ) %>% 
    filter(eligible == 1,
           omicron == 1,
           vaccine_cat %in% c("d2 254+ PF",
                              "d3 7-66 3PF",
                              "d3 67-126 3PF",
                              "d3 127+ 3PF",
                              "d2 194+ AZ",
                              "d3 7-66 2AZ/1PF",
                              "d3 67+ 2AZ/1PF")) %>% 
    mutate(vaccine_cat = fct_drop(vaccine_cat),
           primary_inf = factor(primary_inf, labels = pi_label))

# check vaccine category groupings
model_data %>%
    group_by(vaccine_cat) %>%
    summarise(
        particiants = n_distinct(study_id),
        events = sum(event),
        exposure = sum(tstop - tstart)
    ) %>%
    flextable() %>% 
    autofit()

# primary infection category groupings
model_data %>%
    group_by(primary_inf) %>%
    summarise(
        particiants = n_distinct(study_id),
        events = sum(event),
        exposure = sum(tstop - tstart)
    ) %>%
    flextable() %>% 
    autofit()

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    labs(title = "Kaplan-Meier survival curve")+
    theme_minimal()

# test proportional hazards assumption
# not significant - indicates proportional hazards assumption holds
uni_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

uni_ph$table %>% knitr::kable(caption = "Proportional hazards for univariable model")

multi_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

multi_ph$table %>% knitr::kable(caption = "Proportional hazards for multivariable model")

# table, finalfit provides both unadjusted and adjusted hazards
# model_data %>% finalfit(dependent, explanatory_multi) %>% 
#     flextable() %>% 
#     autofit()

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```

### Hazard ratios from multivariable model

```{r ve-fig-om, fig.height=10, fig.width=15, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```

```{r om-efficacy}

cox_result <- model_data %>% coxphmulti(dependent, explanatory_multi)

# # vaccination
# cox_result %>% 
#     broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
#     filter(grepl("vaccine", term)) %>% 
#     mutate(term = sub("vaccine_cat","",term),
#            term = factor(term, levels = ve_label)) %>% 
#     ggplot() + 
#     aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high, label= round(1-estimate,2)) + 
#     geom_point() +
#     geom_linerange() +
#     geom_text(y = 1)+
#     labs(x = "", y = "1 - HR", title = "Efficacy of vaccination against infection during Omicron period", subtitle = "Reference: Unvaccinated") +
#     coord_cartesian(ylim = c(0,1)) +
#     theme_minimal()+
#     theme(axis.text.x = element_text(angle = 45, hjust=1))

# prior infection
p2 <- cox_result %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
    filter(grepl("primary", term)) %>% 
    mutate(term = sub("primary_inf","",term),
           term = factor(term, levels = pi_label)) %>% 
    ggplot() + 
    aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high) + 
    geom_point(color = "#F8766D") +
    geom_linerange(color = "#F8766D") +
    labs(x = "", y = "1 - HR", subtitle = "A. Fine timescale") +
    coord_cartesian(ylim = c(0,1)) +
    theme_minimal()


# prior infection, 1 year cutoff
model_data <- model_data %>% 
    mutate(primary_inf = fct_collapse(primary_inf, "<12 months" = c("3-6 months","6-9 months","9-12 months"),
                                                   "12+ months" = c("12-15 months","15+ months")))

cox_result <- model_data %>% coxphmulti(dependent, explanatory_multi)

p3 <- cox_result %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>% 
    filter(grepl("primary", term)) %>% 
    mutate(term = sub("primary_inf","",term)) %>% 
    ggplot() + 
    aes(term, 1 - estimate, ymin = 1 - conf.low, ymax = 1 - conf.high) + 
    geom_point(color = "#F8766D") +
    geom_linerange(color = "#F8766D") +
    labs(x = "", y = "1 - HR", subtitle = "B. Course timescale") +
    coord_cartesian(ylim = c(0,1)) +
    theme_minimal()

p2 + p3 + plot_annotation(title = "Efficacy of prior infection against future infection during Omicron period", caption = "Reference group: Naive") + plot_layout(widths = c(2, 1))

```


## Session info

```{r session-info, results='asis'}

sessionInfo() %>%
    pander::pander()

```
