---
title: "SIREN Cox models - R implementation"
author: 
- "Peter Kirwan, MRC Biostatistics Unit, University of Cambridge"
date:   "`r Sys.Date()`"
output: 
    html_document:
        code_folding: hide
        toc:          true
---

```{r, include=FALSE}

# libraries
library(tidyverse)
library(survival)
library(finalfit)
library(flextable)

# load data
load("~/coviddata/siren_post_processing.RData")

```

```{r labels-functions}

# define labels
ve_label <- c(
    "Unvaccinated",
    "Pfizer dose 1, 0-20 days",
    "Pfizer dose 1, >20 days",
    "Pfizer dose 2, 0-13 days",
    "Pfizer dose 2, 14-193 days",
    "Pfizer dose 2, >193 days, Delta",
    "Pfizer dose 3, 0-6 days, Delta",
    "Pfizer dose 3, 7+ days, Delta",
    "ChAdOx + Pfizer dose 3, 0-6 days, Delta",
    "ChAdOx + Pfizer dose 3, 7+ days, Delta",
    "Pfizer dose 2, >193 days, Omicron",
    "Pfizer dose 3, 0-6 days, Omicron",
    "Pfizer dose 3, 6-66 days, Omicron",
    "Pfizer dose 3, 67+ days, Omicron",
    "ChAdOx + Pfizer dose 3, 0-6 days, Omicron",
    "ChAdOx + Pfizer dose 3, 6-66 days, Omicron",
    "ChAdOx + Pfizer dose 3, 67+ days, Omicron"
)

```

This document presents Cox Proportional Hazards estimates for the SIREN interim analysis dataset (SIREN_Interim3_20220104.dta). This analysis replicates a set of Stata analyses, thereby providing independent verification.

Survival analysis datasets with multiple intervals for each participant and corresponding covariate values were prepared using the `survival` package (_Therneau T_) and formatted using the `finalfit` package (_Harrison E, Drake T, and Ots R_).

Minor differences exist between the Stata and `survival` package implementations of the Cox Proportional Hazards model:

- The default method for handling tied event times in Stata is the Breslow approximation, whereas in the `survival` package the more accurate Efron approximation is the default. We decided to use the Efron approximation.

- Clustering on NHS trust is used for robust variance calculation. When using clustering, Stata makes an adjustment to the standard errors - multiplying by g/(g − 1), where g is the number of clusters - whereas the `survival` package does not make this adjustment. We decided to use the adjusted standard errors.

## 1. Naive cohort

The vaccination categories were: unvaccinated, 0/20 days post-first dose, 0/14/194 days post-second dose, and 0/7/67 days post-third dose.

The first two doses considered Pfizer only, with the third dose sub-divided according to the vaccine regimen: 3 x Pfizer, or 2 x ChAdOx + Pfizer, and by period: Delta (pre-December 2021) or Omicron (post-December 2021).

Follow-up time for those within 90 days of a primary infection was removed. The "ChAdOx + Pfizer dose 3, 0-6 days, Omicron" category was removed due to insufficient observations.

```{r ve, fig.width=10, warning=FALSE, message=FALSE}

# define variables
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "vaccine_cat"
explanatory_multi = c("vaccine_cat", "gender", "ethngr", "strata(region)", "strata(agegr)", "strata(work_exposure_frequency)", "strata(occ_set)", "cluster(trust_code)")

# format data
model_data <- split_ve %>% mutate(vaccine_cat=factor(vaccine_cat, labels = ve_label)) %>% 
    filter(eligible == 1, primary_inf==0, vaccine_cat!="ChAdOx + Pfizer dose 3, 0-6 days, Omicron") %>% 
    mutate(vaccine_cat = fct_drop(vaccine_cat))

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    theme_minimal()

# test proportional hazards assumption
# not significant - indicates proportional hazards assumption holds
uni_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

uni_ph$table %>% knitr::kable(caption = "Proportional hazards for univariable model")

multi_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

multi_ph$table %>% knitr::kable(caption = "Proportional hazards for multivariable model")

# table, finalfit provides both unadjusted and adjusted hazards
model_data %>% finalfit(dependent, explanatory_multi) %>% 
    flextable() %>% 
    autofit()

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```

```{r ve-fig, fig.height=10, fig.width=15, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```

## 2. Positive cohort

As above, except considering those who are positive at cohort entry, or become positive prior to first vaccine dose.

The comparator group is kept as unvaccinated naive.

```{r ve-pos, fig.width=10, warning=FALSE, message=FALSE}

ve_label_pos <- c("Unvaccinated, naive", ve_label)

# define variables
dependent = "Surv(tstart, tstop, event)"
explanatory_uni = "vaccine_cat"
explanatory_multi = c("vaccine_cat", "gender", "ethngr", "strata(region)", "strata(agegr)", "strata(work_exposure_frequency)", "strata(occ_set)", "cluster(trust_code)")

# format data
model_data <- split_ve %>% 
        mutate(
            vaccine_cat = if_else(primary_inf==0 & vaccine_cat==0, 0, as.double(vaccine_cat)),
            vaccine_cat = factor(vaccine_cat, labels = ve_label_pos)) %>% 
    filter(
        eligible == 1, 
        (primary_inf==2 | vaccine_cat=="Unvaccinated, naive"), 
        vaccine_cat!="ChAdOx + Pfizer dose 3, 0-6 days, Omicron") %>% 
    mutate(vaccine_cat = fct_drop(vaccine_cat))

# kaplan meier survival curves
survfit(as.formula(paste0(dependent, "~", explanatory_uni)), 
                  data = model_data) %>% 
    broom::tidy() %>% 
    ggplot()+
    aes(time, estimate, color = strata)+
    geom_step()+
    theme_minimal()

# test proportional hazards assumption
# not significant - indicates proportional hazards assumption holds
uni_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_uni) %>% 
    cox.zph()

uni_ph$table %>% knitr::kable(caption = "Proportional hazards for univariable model")

multi_ph <- model_data %>% 
    coxphmulti(dependent, explanatory_multi) %>% 
    cox.zph()

multi_ph$table %>% knitr::kable(caption = "Proportional hazards for multivariable model")

# table, finalfit provides both unadjusted and adjusted hazards
model_data %>% finalfit(dependent, explanatory_multi) %>% 
    flextable() %>% 
    autofit()

# if we wish to match the standard error calculation from Stata:
# sqrt(diag(vcov(siren_cox))* (135/134))

```

```{r ve-fig-pos, fig.height=10, fig.width=15, warning=FALSE, message=FALSE}

# figure
model_data %>% 
    hr_plot(dependent, explanatory_multi, dependent_label = "Survival")

```


1.	Divide in primary infection <1year and >1year.
2.	Probably move Omicron period to 10-15 December 2021 (currently is one).
3.	Create Omicron/Alpha-Delta divide for unvaccinated as well.
4.	Possibly start analysis in December, when Omicron starts.
5.	Include people who enrolled after infection and vaccination in the positive cohort.
6.	Move people who start in the negative cohort and are infected after vaccination to the positive cohort.



## Session info

```{r session-info, results='asis'}

sessionInfo() %>%
    pander::pander()

```
